Clustering Beat-Chroma Patterns\\ in a Large Music Database

### 2010-03-24 Paper/Paper.tex
eddited section 2

\begin{document}
%
\maketitle
%
\begin{abstract}
A musical style or genre implies a set of common conventions
and patterns combined and deployed in different ways to make
individual musical pieces; for instance, most would agree that
contemporary pop music is assembled from a relatively small
palette of harmonic and melodic patterns.  The purpose of this
paper is to use a database of tens of thousands of songs
in combination with a compact representation of melodic-harmonic
content, the beat-synchronous chromagram, and data-mining
tools (clustering) to attempt to explicitly catalog this palette --
at least within the limitations of the beat-chroma representation.
We use online $k$-means clustering to summarize 3.7 million
4-beat bars in a codebook of a few hundred prototypes.
By measuring how accurately such a quantized codebook
can reconstruct the original data, we can quantify the degree
of diversity (distortion as a function of codebook size) and
temporal structure (\ie the advantage gained
by joint quantizing multiple frames) in this music.  The most
popular codewords themselves reveal the common chords
used in the music.  Finally, the quantized representation of
music can be used for music retrieval tasks such as artist
and genre classification, and identifying songs that are
similar in terms of their melodic-harmonic content.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}
The availability of very large collections of music audio present
many interesting research opportunities.  Given millions of examples
from a single, broad class (\eg, contemporary commercial pop music),
can we infer anything about the underlying structure and common
features of this class?  This paper describes our work in this direction.

@@

%\section{Algorithm}\label{sec:algo}
In this section we present the creation of the harmonic patterns from
the Echo Nest data. Then, we explain the online vector quantization algorithm
used to learn a codebook of those patterns.

\subsection{Pattern Creation}
Patterns are defined as a number of beats or bars. In the case of beats,
we simply cut the set of chromas per beat every $N$ beats. In the case
of bars, we gather all beats corresponding to the $N$ bars requested.
The number of chroma vectors we get can vary. We resize the features in the
time axis to a fixed length. Resizing over clipping can be argued over, 
but it should
preserve the main bar characteristics (such as a strong beat at the beginning
or a note change in the middle).

We try to normalize the patterns in respect to the key. We roll the matrix
so the row with the more energy is always the lowest one. This can easily be
seen in Figure \ref{fig:code}. We considered the same techniqueh for rolling
along the other axis and be invariant regarding the downbeat. However, it
seems less robust to noise. Note that we roll each pattern independently.
Also, this implies that the pitch envelope is important rather than the actual 
pitches.

\subsection{Codebook Learning}
We use an online version of the vector quantization algorithm 
\cite{Gersho1991}. This can also be seen as an online K-means.
The idea is as follow: take a sample from your data, find the closest
codeword in the codebook, bring that codeword closer to the 
sample by some amount.
The details are explained in Algorithm \ref{algo:vq}. Note that this
algorithm, though not optimal, is $O(n)$ with $n$ number of songs seen.
Thus it scales linearly to any size of data set. The setting of the learning
rate might become trickier, but this is not fundamentally a problem.

We initialize the codebook by choosing $K$ random points from our dataset.

>>>>

\begin{document}
%
\maketitle
%
\begin{abstract}
A musical style or genre implies a set of common conventions
and patterns combined and deployed in different ways to make
individual musical pieces; for instance, most would agree that
contemporary pop music is assembled from a relatively small
palette of harmonic and melodic patterns.  The purpose of this
paper is to use a database of tens of thousands of songs
in combination with a compact representation of melodic-harmonic
content, the beat-synchronous chromagram, and data-mining
tools (clustering) to attempt to explicitly catalog this palette --
at least within the limitations of the beat-chroma representation.
We use online $k$-means clustering to summarize 3.7 million
4-beat bars in a codebook of a few hundred prototypes.
By measuring how accurately such a quantized codebook
can reconstruct the original data, we can quantify the degree
of diversity (distortion as a function of codebook size) and
temporal structure (\ie the advantage gained
by joint quantizing multiple frames) in this music.  The most
popular codewords themselves reveal the common chords
used in the music.  Finally, the quantized representation of
music can be used for music retrieval tasks such as artist
and genre classification, and identifying songs that are
similar in terms of their melodic-harmonic content.
\end{abstract}
%
\section{Introduction}\label{sec:introduction}
The availability of very large collections of music audio present
many interesting research opportunities.  Given millions of examples
from a single, broad class (\eg contemporary commercial pop music),
can we infer anything about the underlying structure and common
features of this class?  This paper describes our work in this direction.

@@

%\section{Algorithm}\label{sec:algo}
In this section we present the creation of the harmonic patterns from
the Echo Nest data.  Then we explain the online vector quantization algorithm
used to learn a codebook of those patterns.

%\subsection{Harmonic Patterns}
\subsection{Beat-Chroma Patches}
% Patterns are defined as a number of beats or bars. In the case of beats,
% we simply cut the set of chromas per beat every $N$ beats. In the case
% of bars, we gather all beats corresponding to the $N$ bars requested.
% The number of chroma vectors we get can vary. We resize the features in the
% time axis to a fixed length. Resizing over clipping can be argued over, 
% but it should
% preserve the main bar characteristics (such as a strong beat at the beginning
% or a note change in the middle).

We use the bar segmentation obtained from the analysis described in
the previous section to break a song into a collection of beat-chroma
``patches'', typically one or two bars in length. 
%
Because the bar length varies, depending on the meter of a particular
song (see Figure \ref{fig:barsize}), we resample each patch to a fixed
length of typically 4 to 8 beats.
% although the majority of the data contains bars that are 4 beats long

Finally, we normalize the patches with respect to the key by rotating
the pattern matrix so that the first row contains the most
energy. This can easily be seen in Figure \ref{fig:code}.
% We considered the same technique for rolling along the other axis and
% be invariant regarding the downbeat. However, it seems less robust to
% noise.
Each patch within a song is normalized independently, so exact
reconstruction of the original song requires knowledge of the
rotation index (key) for each patch.

The representation resulting from this process is invariant to both
the key and meter of the original song.  This enables the study of
broad harmonic patterns found throughout the data, without regard for
the specific musical context.
%
% In the context of clustering, this enables us to study harmonic
% pattern usage without maintaining clusters for the specific context a
% particular patterns occurs in,
In the context of clustering this avoids \eg obtaining separate
clusters for every major triad in duple or triple meter.
% Also, this implies that the pitch envelope is important rather than
% the actual pitches.
Something about why we're not interested in specific chords?

\subsection{Clustering}

We use an online version of the vector quantization algorithm
\cite{Gersho1991}, to cluster the beat-chroma patches described in the
previous section.
% This can also be seen as an online K-means.
For each sample from the data, the algorithm finds the closest cluster
in the codebook and updates the cluster centroid (codeword) to be
closer to the sample based on some learning rate $l$.
The clusters are updated as each data point is seen, as opposed to
once per iteration in the standard k-means algorithm.
The details are explained in Algorithm \ref{algo:vq}.
%
As in standard k-means clustering, the codebook is initialized by
choosing $K$ random points from our dataset.
%
Note that this algorithm, although not optimal, scales linearly with
the number of patches seen and can be interrupted at any time
while still resulting in an updated codebook. %not that this is important...
% The setting of the learning rate might become trickier, but this is
% not fundamentally a problem.
